---
title: "MB5370 Module 4 Workshop 1"
author: "Nicoline Winding"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible research & version control

## Workshop 1 code and activities
May 2024

Workspace for R code: ~/Documents/github/mb35370-mod4-personal/code/workshop1.R

```{r setting up github, include=FALSE}
#install.packages("usethis")
#introducing myself to GitHub
#library(usethis)
#use_git_config(user.name = "nicolinewinding", user.email = "nicoline.winding@hotmail.com")

##check if everything is working
#install.packages("credentials")
#credentials::git_credential_ask()
#usethis::git_sitrep()

##change credentials
#gitcreds::gitcreds_set()

##get git to talk to github
#usethis::use_github()
## now I can use the pull and push buttons, which just tells me that everything is up-to-date
```

Install and load tidyverse packages
```{r packages}
## install.packages("tidyverse") # Delete this line once installed
library("tidyverse")
```

Look at the data frame
```{r dataframe}
data(mpg)
#if you want to know more about this dataset
?mpg
```

Check that data looks correct
```{r data checks}
head(mpg)
glimpse(mpg)
summary(mpg)
```

## Graphing template

Create first plot
```{r first plot}
#put the size of the engine (displ) on the x axis and the fuel efficiency (hwy) on the y axis
#the function geom_point() adds a layer of points to your plot
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot shows a negative relationship between engine size (displ) and fuel efficiency (hwy). In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?



# Understanding grammar of graphics
What happens if we just try to run ggplot on its own?
```{r}
ggplot() 
# Or with just the data specified?
ggplot(data = mpg)
```

## Aesthetic mappings

```{r aesthetics mapping}
#add a third variable, such as class, by mapping it to an aesthetic represented as aes() in your template from above
#change point colours by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
#change point size by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
#change point transparency by class (alpha)
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
#change point shape by class
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
#change all points to pink
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "pink")
#change colour based on value of one of the variables
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = displ < 5))
```



## Troubleshooting 
ggplot(data = mpg) 
+ geom_point(mapping = aes(x = displ, y = hwy)) 
ERROR: the plus sign should be on the top row of code

## Facet and panel plots
Use this in order to develop separate plots for subsets of your data. In ggplot this is called facets:
```{r facets}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) #this creates 7 separate plots, one for each class
```

You can also create a grid of separate plots for all the variable combinations - but you can only use this for discrete variables:
```{r grid}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

You can also choose to only display coloums, but no rows in the grid of separate plots:
```{r only coloumns}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
```
Exercise: 'nrow' and 'ncol' changes the number of rows and coloums displayed in a grid


## Fitting simple lines

Use geompoint to display data as points:
```{r geompoint}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

Display the same data as a smooth line fit through the points:
```{r geomsmooth}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

You can also change the line type and use a variable to control it
```{r geomsmooth1}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) #changes line type depending on the drv of the car
# changing the colour of each line based on the drv value
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE,)
```

You can also plot multiple geoms on single plot to show underlying data for the smoothed line
```{r multiple geom}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
#there is a more efficient way of doing the same thing
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
#when you code it more efficiently you can change style for the points and lines more easily
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
#you can also map only a subset of the data by using a filter
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

# Exercise:
1.  What geom would you use to draw a line chart? 
  A boxplot? geom_boxplot()
  A histogram? geom_histogram()
  An area chart? geom_area()
2.  Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
  I think they will look the exact same
```{r exercise 1}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

### Transformations and Stats
Here we will learn about easy transformations and data summaries

## Plotting statistics
Plot a bar chart to show availability of different cut qualities
```{r diamonds}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut)) #the x axis displays a variable not in the data set, this is because geom_bar groups your data into counts by calculating statistical transformations

#you can make the same exact figure by using a stat command instead
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #because every geom has a default stat
```
## Overriding defaults in ggplot2
Change the default stat to an identity

```{r new data}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity") #this overrides the default stat
```

You can also override a default mapping from transformed variables to aesthetics
```{r override default}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1)) #this shows proportion rather than count on the x axis
```

## Plotting statistical details
Showing uncertainty or limitations is good practice, and this can be done by using stat_summary()
```{r stat_summary }
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

### Aesthetic adjustments

Colour and fill can also convey a lot of information
```{r color and fill}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

You can also use aesthetics on specific variables
```{r color and fill2}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity)) #this uses the position argument automatically
```

Using 'position = "identity" will allow you to place each object excatly where it falls in the context of the graph
```{r color and fill3}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") #looks great

ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity") #bit confusing
```

Using 'position = "fill" makes each stacked bar the same height and shows data as proportion of 1 instead
```{r color and fill4}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

Using 'position = "dodge" places overlapping objects beside each other rather than stacked
```{r color and fill5}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

Using 'position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap
```{r color and fill6}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```






