---
title: "Workshop4"
output: html_document
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial data in R

## Workshop 4 code and activities
May 2024

R can do practically everything that the off-the-shelf GIS programs can do, with some added benefits:
- you don't need to remember complex workflows based on mouse clicks and maintaining processing logs
- you can version control and backup your spatial analysis, and make you work reproducible and openly avaialable
- you can interface your results directly with other components of the R system

## Installing the spatial R packages
```{r packages, include=FALSE}
#install.packages("sf") 
#install.packages("terra")
#install.packages("tmap")


#load into R library
library(tidyverse)
library(sf) # simple features
library (terra) # for raster
library(tmap) # Thematic maps are geographical maps in which spatial data distributions are visualized DOESN'T WORK!!!!!!!
```

```{r load data, include=FALSE}
library(readr)
dat <- read_csv("../data/workshop4/copepods_raw.csv")
dat
```

## Data exploration
It is not uncommon to be given data by collaborators you know nothing about, so always good practice to thoroughly check it before running any code on it.

### Check coordinates
```{r check coords, include=FALSE}
library(ggplot2)
ggplot(dat) + 
  aes(x = longitude, y = latitude, color = richness_raw) +
  geom_point()
```

This shows the location of every segment as well as the species richness depicted by the color of the points. This is useful but not a map.

We can also visualise species richness in a non-spatial way:
```{r non spatial visualisation, include=FALSE}
ggplot(dat, aes(x = latitude, y = richness_raw)) + 
  stat_smooth() + 
  geom_point()
```
This looks odd with data points at latitudes above -40. We can try and look at other variables and talk to collaborators to figure out what might be causing these weird patterns.


## Getting going with maps
We can use some specialist R packages to reference data to real locations on the planet and bend it around a mostly spherical ball that is the earth.

This turns our data into a simple features collection:
```{r using sf, include=FALSE}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
```

## Coordinate reference systems

You need coordinate reference systems for 2D mapping to compensate for lumpy 3D nature of the earth.
```{r using sf2, include=FALSE}
crs4326 <- st_crs(4326)
crs4326 # look at the whole CRS
crs4326$Name # pull out just the name of the crs
```

## Feature collection (points)

Let's look at what we created with sdat:
```{r sdat, include=FALSE}
sdat
```

A sf is like a shapefile that holds data in columns and rows but is also spatially aware by including extra columns regarding each rows position and metadata about the coordinate reference system, the type of geometry, etc.

## Cartography

Let's only plot the richness column:
```{r sdat plot, include=FALSE}
plot(sdat["richness_raw"])

#this plots all the columns:
plot(sdat)
```

## Thematic maps for communication

The tmap package is similar to ggplot as we build maps and add layers.
tm_dots plots dots of the coordinates and we choose "richness:raw" as the color scale:
```{r using tmap, include=FALSE}
tm1 <- tm_shape(sdat) + 
  tm_dots(col = "richness_raw")

tm1
```

You can also save these maps with the command: tmap_save, similarly to ggsave.

## Mapping spatial polygons as layers

The sf package can handle many types of spatial data, including shapes and polygons.

### Loading shapefiles

.shp files are generally considered an undesirable file format because they are inefficient at storing data and to save one shapefile you actually create multiple files, which means bits of the file can be lost if you transger your data around. It is better to move away from shapefiles and save files as .gpkg

```{r load files, include=FALSE}
aus <- st_read("../data/workshop4/spatial-data/Aussie/Aussie.shp")

shelf <- st_read("../data/workshop4/spatial-data/aus_shelf/aus_shelf.shp")

#check data
aus
shelf
```
### Mapping your polygons

```{r map polygon, include=FALSE}
tm2 <- tm_shape(shelf) + 
  tm_polygons()

tm2
```

We can add layers to make thematic maps:
```{r add layers, include=FALSE}
tm3 <- tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots() 

tm3
```
## Exploring t_map

Now we will customize our map:

```{r customize map, include=FALSE}
?tmap

vignette('tmap-getstarted')

#install.packages("viridisLite")
library(viridis)

tmap_mode("view") #makes the map in the following code interactive in the sense you can zoom in and out on the whole world

tm4 <- tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) +
  tm_dots(col = "richness_raw", palette ="-viridis") +
  tmap_style("albatross") +
  tm_layout(title = "Richness", bg.color = "#228B22") +
  tm_scale_bar(position = c("right", "bottom"), width = 0.15) #+
  #tm_compass(position = c("left", "bottom"), size = 2) this doesn't work with the interactive view mode

tm4
```

There are also heaps of commands to insert different maps in the same plot, such as grid.layout and pushViewport and tm_facets.

There is also an option to make maps animated with commands such as tmap_animation and loop, but these of course cannot be used in publications.

## Exporting your map

```{r saving map, include=FALSE}

tmap_save(tm4, filename = "../output/Map.jpg", 
          width = 150, height = 150, units = "mm") #obviously cannot save the map as interactive
```

You can also import raster data to intersect with point data to create even more informative maps!

