---
title: "Workshop2"
output: html_document
date: "2024-05-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using ggplot2 for communication

## Workshop 2 code and activities


## Labels

Labels can help you communicate your key findings and comply with formatting requirements of any outlet.

```{r labels, include=FALSE}
#create an nice title
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se.e = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")

#subtitle and caption
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )
```

The same can be achieved by using the command labs()
```{r labs, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

## Annotations
We can also add text to the plot itself by using things like geom_text() 

```{r filter and annotation, include=FALSE}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class) 
#these overlap a bit but you can use functions such as nudge() and other packages to deal with this
```

## Scales

You can change the default scales of ggplot to costumise plots and improve communication.
```{r scales, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous(limits = c(0,10)) +
  scale_y_continuous(limits = c(20,30)) +
  scale_colour_discrete()
```

## Axis ticks

You can also change the ticks on the axis by using labels as an argument to change the text label associated with ticks:

```{r ticks, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) #this changes the ticks on the y axis to increase by 5 at a time

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) #this removes the labels of all the ticks
```

## Legends and colour schemes

You can change the position of your legend to make more room for the plot or even hide it altogether

```{r legend, include=FALSE}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") #the ggplot default
base + theme(legend.position = "none") #completely removes legend
```

## Replacing a scale
2 types of scales you are most likely to want to switch out: continuous and colour scales. 

We can also log transform data to scale the data more precisely:
```{r log scale, include=FALSE}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

We can also change the colour scale to communicate data more effectively:
```{r colour scale, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")

#you can also change shape to make sure data is interpretable in B&W
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")

#you can also use predefined colours
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

There are also packages that allow you to change colours
```{r colour packages, include=FALSE}
#install.packages('viridis')
#install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

## Themes

You can also costumise and entire theme of a plot. ggplot has 8 default themes and you can use packages to use more themes

```{r themes, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw() #white background with black border

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light() #white background and grey border

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic() #no gridlines or border, just black x and y axis

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark() #dark grey background and no border
```

You can also develop you own theme by setting some or all arguments in theme() yourself. Nick's example:
```{r Nick's theme, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```

## Saving and exporting your plots
You can save your work using ggsave() rather than exporting the plot window. This allows you to costumise the settings.


